Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]

### 2020-08-05 paper/manuscript.tex
Submission version.

\newcommand{\versionnumber}{0.1} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations: v\versionnumber}
\author[1*]{Antonia S. J. S. Mey}
\author[7]{Bryce K. Allen}
\author[2]{Hannah E. Bruce Macdonald}
\author[2*]{John D. Chodera}
\author[1,10]{Maximilian Kuhn}
\author[1*]{Julien Michel}
\author[3*]{David L. Mobley}
\author[11]{Levi N. Naden}
\author[4]{Samarjeet Prasad}
\author[2,8]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[6*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[7]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[4]{National Institutes of Health, Bethesda, MD, USA}
\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
\affil[7]{Silicon Therapeutics, Boston, MA, USA}
\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Molecular Sciences Software Institute, Blacksburg VA, USA}
%
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{mail@julienmichel.net}{JM}
\corr{michael.shirts@colorado.edu}{MRS}
%
\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Month, Day, Year}
\dateaccepted{Month, Day, Year}

@@

\paragraph{Protonation states depend on the pH of the experimental assay}
Care should be taken when preparing ligands and proteins to match the pH of the experimental assay, if known. As mentioned above in Sec.~\ref{subsec:exp_condition}, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands. Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improving, 2020schrodinger}. Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionised, but care is still needed as the local environment can modify expected ionization states (for instance the catalytic Asp dyad in proteases). Histidine is notoriously more difficult to predict as its pKa suggests it ionizes closer to the experimental pH range. For ligands often the pKa needs to be determined, if not known experimentally. There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}. Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools~\cite{isik2018pka}. While often it can be assumed that the protonation state of a ligand and protein will remain the same as a ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}. BACE, for example, famously undergoes a protonation state change on ligand binding.

@@

For binding, however, the situation is more complex and requires special care.
Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
\begin{equation}\label{eq:restraint-correction}
    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( \frac{c^{\circ}}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right)
\end{equation}
where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
This is typically hard to achieve as the exploration process is governed by diffusion.
The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
When restraint are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.

@@

\paragraph{Handling multiple potential binding modes within relative calculations.}
Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy~\cite{}.

@@

As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally -determined affinity values, plots should be depicted with the same range on both axes (i.e. $x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same. If this skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g. 1 cm is 1 kcal/mol is acceptable. Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions. These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable (and handled as outliers). In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas.

@@

\begin{checklist}{Running absolute simultions}
        \textbf{Steps to running your absolute alchemical free energy calculations}
\begin{itemize}
 \item Check your ligands have the same, biologically correct binding pose
        \item Make sure your \textlambda-scheduling is set appropriately
        \item Check if your ligands are discharging and vanishing correctly
        \item Set up your restraints correctly
        \item Make sure you subsample the data in your free energy estimation protocol
        \item Apply the appropriate correction terms
\end{itemize}
\end{checklist}

\begin{checklist}{Running relative simulations}
        \textbf{Steps to running your relative alchemical free energy calculations}
\begin{itemize}
   \item Check your ligands  have the same, biologically correct binding pose
        \item Make sure your λ-scheduling is set correctly
        \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures to check consistency in the results
        \item Check whether dummy atoms were assigned correctly
        \item Make sure you subsample the data in your free energy estimation protocol
        \item Apply the 
        appropriate correction terms
\end{itemize}
\end{checklist}
\end{Checklists*}

%Analyis checklist
\begin{Checklists*}
\begin{checklist}{How do I know which simulations are unreliable?}
    \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
        \begin{itemize}
                \item Standard error (\textsigma) should not be \textgreater1 kcal·mol$^{-1}$ 
    \item Simulated systems have not converged - this
    \newline\newline\textit{Relative:}
    \item If you observe hysteresis in perturbations and incorrect cycle closures
    \item Energy differences \textgreater15 kcal$\cdot$mol$^{-1}$  are unreliable
    \newline\newline\textit{Absolute:}
    \item Energies below -15 kcal$\cdot$mol$^{-1}$  are unreliable
    \item The ligand has not sampled most of the intended region after the vanishing step
    \item The ligand is drifting out of the intended region after the vanishing step
        \end{itemize}
\end{checklist}

@@

\begin{checklist}{Data Analysis}
    \textbf{Steps to analyzing your output data correctly}
\begin{itemize}
    \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
    \item You should be able to explain large energy differences based on your system (trajectory)
\end{itemize}
\end{checklist}
\end{Checklists*}
\clearpage

@@

\section*{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
ASJSM and JM acknowledge funding through an EPSRC flagship software grant: EP/P022138/1
MK and JM acknowledge funding through Innovate UK by KTP partnership 011120.
AR acknowledges partial support from the Tri-Institutional Program in Computational Biology and Medicine and the Sloan Kettering Institute.
HEBM acknowledges support from a Molecular Sciences Software Institute Investment Fellowship and Relay Therapeutics.
JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories. A complete funding history for the Chodera lab can be found at http://choderalab.org/funding. DLM appreciates financial support from the National Institutes of Health (1R01GM108889-01, 1R01GM124270-01A1 and R01GM132386), and the National Science Foundation (CHE 1352608).

\bibliography{alchemical,manual}

>>>>

\newcommand{\versionnumber}{1.0} % you should update the minor version number in preprints and major version number of submissions.
\newcommand{\githubrepository}{\url{https://github.com/michellab/alchemical-best-practices}} %this should be the main github repository for this article
\newcommand{\expect}[1]{\left\langle{#1}\right\rangle}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% ARTICLE SETUP
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Best Practices for Alchemical Free Energy Calculations [Article v \versionnumber]}
\author[1*]{Antonia S. J. S. Mey}
\author[7]{Bryce K. Allen}
\author[2]{Hannah E. Bruce Macdonald}
\author[2*]{John D. Chodera}
\author[1,10]{Maximilian Kuhn}
\author[1]{Julien Michel}
\author[3*]{David L. Mobley}
\author[11]{Levi N. Naden}
\author[4]{Samarjeet Prasad}
\author[2,8]{Andrea Rizzi}
\author[1]{Jenke Scheen}
\author[6*]{Michael R. Shirts}
\author[9]{Gary Tresadern}
\author[7]{Huafeng Xu}
%
\affil[1]{EaStCHEM School of Chemistry, David Brewster Road, Joseph Black Building, The King's Buildings, Edinburgh, EH9 3FJ, UK}
\affil[2]{Computational and Systems Biology Program, Sloan Kettering Institute, Memorial Sloan Kettering Cancer Center, New York NY, USA}
\affil[3]{Departments of Pharmaceutical Sciences and Chemistry, University of California, Irvine, USA}
\affil[4]{National Institutes of Health, Bethesda, MD, USA}
\affil[6]{University of Colorado Boulder, Boulder, CO, USA}
\affil[7]{Silicon Therapeutics, Boston, MA, USA}
\affil[8]{Tri-Institutional Training Program in Computational Biology and Medicine, New York, NY, USA}
\affil[9]{Computational Chemistry, Janssen Research \& Development, Turnhoutseweg 30, Beerse B-2340,Belgium}
\affil[10]{Cresset, Cambridgeshire, UK}
\affil[11]{Molecular Sciences Software Institute, Blacksburg VA, USA}
%
\corr{antonia.mey@ed.ac.uk}{ASJSM}
\corr{john.chodera@choderalab.org}{JDC}
\corr{dmobley@mobleylab.org}{DLM}
\corr{michael.shirts@colorado.edu}{MRS}

\orcid{Antonia S. J. S. Mey}{0000-0001-7512-5252}
\orcid{Bryce Allen}{0000-0002-0804-8127}
\orcid{Hannah E. Bruce Macdonald}{0000-0002-5562-6866}
\orcid{John D. Chodera}{0000-0003-0542-119X}
\orcid{Maximilian Kuhn}{0000-0002-2811-3934}
\orcid{Julien Michel}{0000-0003-0360-1760}
\orcid{David L. Mobley}{0000-0002-1083-5533}
\orcid{Levi N. Naden}{0000-0002-3692-5027}
\orcid{Samarjeet Prasad}{0000-0001-8320-6482}
\orcid{Andrea Rizzi}{0000-0001-7693-2013}
\orcid{Jenke Scheen}{0000-0001-9781-0445}
\orcid{Michael R. Shirts}{0000-0003-3249-1097}
\orcid{Gary Tresadern}{0000-0002-4801-1644}
\orcid{Huafeng Xu}{0000-0001-5447-0452}

\blurb{This LiveCoMS document is maintained online on GitHub at \githubrepository; to provide feedback, suggestions, or help improve it, please visit the GitHub repository and participate via the issue tracker.}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%% PUBLICATION INFORMATION
%%% Fill out these parameters when available
%%% These are used when the "pubversion" option is invoked
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\pubDOI{10.XXXX/YYYYYYY}
\pubvolume{<volume>}
\pubyear{<year>}
\articlenum{<number>}
\datereceived{Month, Day, Year}
\dateaccepted{Month, Day, Year}

@@

\paragraph{Protonation states depend on the pH of the experimental assay}
Care should be taken when preparing ligands and proteins to match the pH of the experimental assay, if known. As mentioned above in Sec.~\ref{subsec:exp_condition}, the pH of the assay can differ from neutral pH and will determine the protonation states of the proteins and ligands. Since the pKa of reference amino acid sidechain residues is known, but can vary in the protein environment, many different tools have emerged for predicting sidechain pKa in proteins, such as the H++ server, ProPKa, APBS, and Maestro ~\cite{anandakrishnan2012automating, sondergaard2011improved, jurrus2018improvements, 2020schrodinger}. Strongly acidic (Glu, Asp) or basic (Arg, Lys) sidechains can reliably predicted to be ionised, but care is still needed as the local environment can modify expected ionization states (for instance the catalytic Asp dyad in proteases). Histidine is notoriously more difficult to predict as its pKa suggests it ionizes closer to the experimental pH range. For ligands often the pKa needs to be determined, if not known experimentally. There are many different available tools for this purpose, but common choices may be propKa~\cite{olsson2011propka3,sondergaard2011improved}, Chemicalize (\url{https://chemicalize.com/welcome}), or Maestro~\cite{2020schrodinger}. Still, accurate pKa prediction for small molecules remains a challenging problem, even with dedicated tools~\cite{isik2018pka}. While often it can be assumed that the protonation state of a ligand and protein will remain the same as a ligand binds, some care needs to be taken with systems where the protonation state may change upon binding~\cite{onufriev2013protonation}. BACE, for example, famously undergoes a protonation state change on ligand binding.

@@

For binding, however, the situation is more complex and requires special care.
Because the simulations are typically performed using restraints and at concentrations that are different from 1 M, the expression of the free energy requires the following correction~\cite{gilson1997statisticalthermodynamic} (see an example of such a thermodynamic cycle in Fig.~\ref{fig:fig_absolute_thermodynamic_cycle})
\begin{equation}\label{eq:restraint-correction}
    \Delta G^{\circ}_{\mathrm{restr}} = -k_BT ~ \ln \left( \frac{c^{\circ}}{V_L} \right) -k_BT ~ \ln \left( \frac{\xi_L}{8 \pi^2} \right),
\end{equation}
where $V_L$ and $\xi_L$ are respectively the volume of the translational and rotational degrees of freedom of the non-interacting ligand in the simulation box.
When no restraints are used, the non-interacting ligand is free to translate and rotate in the simulation box (i.e., $V_L = V_{\mathrm{box}}$ and $\xi_L = 8\pi^2$), and the rotational term is zero.
A sufficiently thorough exploration of the simulation box by the non-interacting ligand is, however, required for the formula to be valid.
This is typically hard to achieve as the exploration process is governed by diffusion.
The addition of a restraint limits the volume available to the non-interacting ligand, thus speeding the convergence of the sampling.
In addition, when enhanced sampling methods such as Hamiltonian $\lambda$ exchange are used (see Sec.~\ref{sec:sampling_schemes}), the use of a restraint is typically necessary as it keeps the ligand in the binding site in the interacting state (see also Sec.~\ref{sec:weak-binders}) and generally reduce the round-trip time of replicas.
When restraint are employed, the values of $V_L$ and $\xi_L$ are restraint-dependent, but for commonly employed restraints, these can be usually easily computed analytically or numerically by solving the relevant integral.

@@

\paragraph{Handling multiple potential binding modes within relative calculations.}
Multiple potential binding modes pose particular problems for relative free energy calculations, as having multiple starting structures for these calculations could yield substantially different calculated relative binding free energies for the same transformation due to kinetic trapping, and, without additional information (specifically, the free energy of binding mode interconversion or, equivalently, the relative populations of different binding modes) it becomes impossible to sort out which of the multiple answers is in fact the correct relative binding free energy.

@@

As we are interested mainly in the linear relationship between the alchemical free energy predictions and the experimentally -determined affinity values, plots should be depicted with the same range on both axes (i.e. $x=y$) with a 1:1 aspect ratio, with units for both experiment and simulation converted to be the same. If this skews the plot to a point where it is difficult to read of information, using the same dimensions, such that e.g. 1 cm is 1 kcal/mol is acceptable. Furthermore, bounds should be depicted for the 1- and 2-kcal/mol confidence regions. These regions can serve as tools to communicate your model performance: any predictions inside the 1 kcal/mol region can be seen as highly reliable, any predictions inside the 2 kcal/mol region should be seen as somewhat reliable, and any predictions outside the confidence regions should be expected to be unreliable (and handled as outliers). In a drug discovery context, this type of data depiction may suggest the reliability of alchemical FE predictions in the project, and can give an idea of how trustworthy predictions can be for synthesis ideas. It is also recommended to included experimental error bars in all plots.

@@

\begin{checklist}{Running absolute simulations}
        \textbf{Steps to running your absolute alchemical free energy calculations}
\begin{itemize}
 \item Check your ligands have the same, biologically correct binding pose
        \item Make sure your \textlambda-scheduling is set appropriately
        \item Check if your ligands are discharging and decoupling correctly
        \item Set up your restraints correctly
        \item Make sure you subsample the data in your free energy estimation protocol
        \item Apply the appropriate correction terms
\end{itemize}
\end{checklist}

\begin{checklist}{Running relative simulations}
        \textbf{Steps to running your relative alchemical free energy calculations}
\begin{itemize}
   \item Check your ligands  have the same, biologically correct binding pose
        \item Make sure your λ-scheduling is set correctly
        \item Make sure your molecular transformations are realistic (1-5 heavy atoms for reliable computations)
        \item Generate a perturbation network by your method of choice; check whether you have enough cycle closures to check consistency in the results
        \item Check whether dummy atoms were assigned correctly
        \item Consider subsampling the data in your free energy estimation protocol
        \item Apply the 
        appropriate correction terms
\end{itemize}
\end{checklist}
\end{Checklists*}

%Analyis checklist
\begin{Checklists*}
\begin{checklist}{How do I know which simulations are unreliable?}
    \textbf{Situations suggesting your relative alchemical free energy calculations have not run properly (assuming absence of experimental affinities)}
        \begin{itemize}
                \item Standard error (\textsigma) should not be \textgreater1 kcal·mol$^{-1}$ 
    \item Simulated systems have not converged - trajectories should be manually checked for consistency; other methods such as generating RMSD plots are also recommended
    \newline\newline\textit{Relative:}
    \item If you observe hysteresis in perturbations and incorrect cycle closures
    \item Energy differences \textgreater$\sim$15 kcal$\cdot$mol$^{-1}$  are likely unreliable
    \newline\newline\textit{Absolute:}
    \item Energies \textless$\sim$-15 kcal$\cdot$mol$^{-1}$  are likely unreliable
    \item The ligand has not sampled most of the intended region after the decoupling step
    \item The ligand is drifting out of the intended region after the decoupling step
        \end{itemize}
\end{checklist}

@@

\begin{checklist}{Data Analysis}
    \textbf{Steps to analyzing your output data correctly}
\begin{itemize}
    \item Make sure you have run enough replicates to ensure statistical reliability (\textgreater3)
    \item Compute both correlation and ranking coefficients and ranking statistics (e.g. r, \textrho, MUE and \texttau)
    \item Include error bars in all your visual analyses
\end{itemize}
\end{checklist}
\end{Checklists*}
\clearpage

@@

\section*{Funding Information}
%%%%%%%
% Authors should acknowledge funding sources here. Reference specific grants.
%%%%%%%
ASJSM and JM acknowledge funding through an EPSRC flagship software grant: EP/P022138/1
MK and JM acknowledge funding through Innovate UK by KTP partnership 011120.
AR acknowledges partial support from the Tri-Institutional Program in Computational Biology and Medicine and the Sloan Kettering Institute.
HEBM acknowledges support from a Molecular Sciences Software Institute Investment Fellowship and Relay Therapeutics.
JDC is a current member of the Scientific Advisory Board of OpenEye Scientific Software and a consultant to Foresite Laboratories. A complete funding history for the Chodera lab can be found at \url{http://choderalab.org/funding}. DLM appreciates financial support from the National Institutes of Health (1R01GM108889-01, 1R01GM124270-01A1 and R01GM132386), and the National Science Foundation (CHE 1352608).

\section*{Author Information}
\makeorcid
\bibliography{alchemical,manual}
